\textls[-0]{\thetitle}

### 2023-04-02 coqimpl.tex
add a bit from rebuttal
address reviewer comments

Importantly,
\lsti{STLC\_tm} merely declares the existence of these names and their types;
it does not specify their definitions.
Having these names and their types available through the context parameters (\lsti{self})
suffices for the translations of the subsequent fields to be type-checked by Coq.
Leaving the definitions unspecified enables \lsti{STLC} and \lsti{STLCFix}
to instantiate \lsti{tm} differently upon \lsti{End}~\lsti{STLC} and upon \lsti{End}~\lsti{STLCFix}.

@@

Rather than modifying the Coq kernel to extend its core theory, a translation to
Coq conveniently reduces the trusted base of any development using \Lang to Coq.
%and improves compatibility with different versions of Coq.
In particular, once a family is closed, \lsti{Print Assumptions} can be used
to verify that there are no lingering \lsti{Axiom}s introduced by the translation.
Ramifications of possible bugs in the \Lang implementation are limited to
the usability of the plugin.
%Finally, we emphasize that our compilation can make sure the trusted code base is restricted to the Coq's kernel, as all the feature supported by our plugin has been compiled into Coq's surface syntax/Vernacular commands.

@@

The \Lang implementation currently does not yet bring extensibility to Coq's full
facility for inductive types.
Mutually inductive types and parameterized inductive types are not yet extensible,
though indexed inductive types are supported and can be used to encode parameterized ones.
Partial recursors are automatically generated only for non-indexed inductive types.
These features are not exercised by our case studies (\cref{sec:casestudies})
but may be useful for modeling other languages.
We believe that they do not pose conceptual challenges and can be
addressed with more engineering effort of the same kind that went into \Lang.

>>>>

Importantly,
\lsti{STLC\_tm} merely declares the existence of these names and their types;
it does not specify their definitions.
Having these names and their types available through the context parameters (\lsti{self})
suffices for the translations of the subsequent fields to be type-checked by Coq.
Leaving the definitions unspecified enables \lsti{STLC} and \lsti{STLCFix}
to instantiate \lsti{tm} differently upon \lsti{End}~\lsti{STLC} and upon \lsti{End}~\lsti{STLCFix}.
In particular, non-exhaustive pattern matching is prevented because
an ordinary recursor like \lsti{tm_rect} is not available.

@@

Rather than modifying the Coq kernel to extend its core theory, a translation to
Coq conveniently reduces the trusted base of any development using \Lang to Coq.
%and improves compatibility with different versions of Coq.
In particular, once a family is closed, \lsti{Print Assumptions} can be used
to verify that there are no lingering \lsti{Axiom}s introduced by the translation.
The ramifications of possible bugs in the \Lang implementation are limited to
the usability of the plugin.
%Finally, we emphasize that our compilation can make sure the trusted code base is restricted to the Coq's kernel, as all the feature supported by our plugin has been compiled into Coq's surface syntax/Vernacular commands.

@@

The \Lang implementation currently does not yet bring extensibility to Coq's full
facility for inductive types.
Mutually inductive types and parameterized inductive types are not yet extensible,
though indexed inductive types are supported and can be used to encode parameterized ones.
Partial recursors are automatically generated only for non-indexed inductive types.
These features are not exercised by our case studies (\cref{sec:casestudies})
but may be useful for modeling other languages.
We believe that they do not pose conceptual challenges and can be
addressed with more engineering effort on the same level as the current \Lang implementation.

### 2023-04-02 metatheory-condensed.tex
address reviewer comments

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
We use based path induction%
\defcitealias{program2013homotopy}{UFP~2013}%
as the elimination principle for identity types~\citepalias{program2013homotopy}.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

We use universes à la \citet{coquand2013presheaf}, following \citet{kaposi2019gluing}.
Unlike Russell-style ones, these universes are not inhabited by types directly,
but rather by the \emph{codes} of types,
and arguably behave better due to its closeness to Tarski-style universes~\cite{luo2012notes}. 
% and compatibility with our algebraic formulation of MLTT~\cite{coquand2018canonicity}. % I am not sure if I want this sentence or not
Term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation to avoid clutter.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

@@

W-types are eliminated with the form $\Wrec{\wsig}{\lkg}{t}$, where $t$ is of a
W-type $\El{\wcode{\wsig}}$, and $\lkg$ is essentially an $n$-tuple of
case handlers for the $n$ constructors in~$\wsig$ (\ruleref{tm/wrec}).
Each case handler has a type of form $\CaseSig{A}{B}{T}$, where
$T$ is the motive of the recursion (\ruleref{tyeq/casety});
for simplicity, we model only non-dependent motives.
The collection of case handlers~$\lkg$ encodes those defined and inherited
by an \lsti{FRecursion} command in our plugin.
We choose to type it with a linkage type $\TyLkg{\RecSig{\wsig}{T}}$
to avoid introducing non-dependent $n$-tuples, which linkages generalize.

@@

Rules \ruleref{tm/wsup} and \ruleref{tm/wrec} require access to~$\wsig$:
the W-type is exhaustively generated by its constructors, and
its elimination must exhaustively handle all the constructors in
its signature.

@@

Derived families can be modeled as linkage transformers inductively
constructed from the introduction forms
$\small\InhId$, $\small\InhExt{h}{\mathit{self}.t}$, $\small\InhOv{h}{\mathit{self}.t}$, %$\small\InhInh{h}$,
etc.
\cref{fig:fmltt-selected} shows the $\beta$-rule of an exemplary transformer,
\ruleref{tmeq/ov/beta}.
It states that applying the transformer $\small\InhOv{h}{\mathit{self}_{\!2}.t_2}$ to
a linkage of form $\LkgAdd{\lkg}{\mathit{self}_{\!1}.t_1}$
overrides the linkage's last field~$t_1$ with~$t_2$.
For instance, the construction below shows that 
$\small\InhOv{\InhId}{\mathit{self}'\!.\wcode{\wsig'_{\texttt{tm}}}}$
is used as the first step in creating a linkage transformer modeling a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$:
\vspace{.9ex}

>>>>

\defcitealias{program2013homotopy}{UFP~2013}

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
We use based path induction
as the elimination principle for identity types~\citepalias{program2013homotopy}.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

We use universes à la \citet{coquand2013presheaf}, following \citet{kaposi2019gluing}.
Unlike Russell-style ones, these universes are not inhabited by types directly,
but rather by the \emph{codes} of types,
and arguably behave better due to its closeness to Tarski-style universes~\cite{luo2012notes}. 
% and compatibility with our algebraic formulation of MLTT~\cite{coquand2018canonicity}. % I am not sure if I want this sentence or not
The term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
The type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation to avoid clutter.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

@@

W-types are eliminated with the form $\Wrec{\wsig}{\lkg}{t}$, where $t$ is of a
W-type $\El{\wcode{\wsig}}$, and $\lkg$ is essentially an $n$-tuple of
case handlers for the $n$ constructors in~$\wsig$ (\ruleref{tm/wrec}).
Each case handler has a type of the form $\CaseSig{A}{B}{T}$, where
$T$ is the motive of the recursion (\ruleref{tyeq/casety});
for simplicity, we model only non-dependent motives.
The collection of case handlers~$\lkg$ encodes those defined and inherited
by an \lsti{FRecursion} command in our plugin.
We choose to type it with a linkage type $\TyLkg{\RecSig{\wsig}{T}}$
to avoid introducing non-dependent $n$-tuples, which linkages generalize.

@@

The rules \ruleref{tm/wsup} and \ruleref{tm/wrec} require access to~$\wsig$:
the W-type is exhaustively generated by its constructors, and
its elimination must exhaustively handle all the constructors in
its signature.

@@

Derived families can be modeled as linkage transformers inductively
constructed from the introduction forms
$\small\InhId$, $\small\InhExt{h}{\mathit{self}.t}$, $\small\InhOv{h}{\mathit{self}.t}$, %$\small\InhInh{h}$,
etc.
\cref{fig:fmltt-selected} shows the $\beta$-rule of an example transformer,
\ruleref{tmeq/ov/beta}.
It states that applying the transformer $\small\InhOv{h}{\mathit{self}_{\!2}.t_2}$ to
a linkage of form $\LkgAdd{\lkg}{\mathit{self}_{\!1}.t_1}$
overrides the linkage's last field~$t_1$ with~$t_2$.
For instance, the construction below shows that 
$\small\InhOv{\InhId}{\mathit{self}'\!.\wcode{\wsig'_{\texttt{tm}}}}$
is used as the first step in creating a linkage transformer modeling a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$:
\vspace{.9ex}
